---
title: "SHMI Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SHMI Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

The Soil Health Management Index (SHMI) turns a record of field management
into a single 0-100 score. It combines four sub-indices, each also scored
0-100, with weights calibrated against measured soil health:

- **Cover** (weight 0.40): the share of days with living plants in the
  growing and non-growing seasons, set by the site's climate.
- **Organic inputs (OrgInput)** (0.32): the share of years with an organic
  amendment or grazing animals.
- **Diversity** (0.15): the average number of plant species per year.
- **Inverse disturbance (InvDist)** (0.13): how little the soil was tilled;
  100 means no tillage.

This vignette walks through the standard workflow: fill in the Excel
workbook, prepare the inputs, add climate, compute SHMI, and review the
results.

SHMI 1.0 was calibrated on 354 US plots at 74 sites of the North American
Project to Evaluate Soil Health Measurements (NAPESHM), mostly annual cropping
systems plus perennial and native reference systems. It is most reliable for
comparing practices at the same site, and should be recalibrated before use
in other regions.

```{r load}
library(SHMI)
```

## 1. Get a workbook

SHMI reads a standard Excel workbook. Start from the blank template, or from
a completed example to see how entries should look:

```{r template, eval = FALSE}
download_shmi_template("SHMI_template.xlsx")  # blank
download_shmi_example()                       # completed example
```

The completed example is also installed with the package, and
`get_shmi_example()` returns its path. The rest of this vignette uses it: eight
management units from a long-term experiment (2020-2024), with continuous
corn and a rotation under several nitrogen and manure treatments.

The workbook has one sheet per kind of record, each with column names on the
fourth row:

| Sheet | Records |
|---|---|
| `Mgt_Unit` | Management units (`MGT_combo`), with study, farm, field, treatment, and optionally the soil sampling date (`MGT_sample_date`) |
| `Crop_Diversity` | Crops: name, category, and plant, harvest, and termination dates |
| `Soil_Disturbance` | Tillage and other soil-disturbing passes |
| `Soil_Amendments` | Amendments, including organic inputs |
| `Animal_Diversity` | Periods with grazing or other animals |

The disturbance, amendment, and animal sheets may be left empty. Each crop is
entered by name with its dates; mixtures are entered as one row per species.
Sequence numbers (`CD_seq_num`) and mixture flags (`CD_mix`) are not needed.

The crop category (`CD_cat`) decides how a crop ends, so it is worth getting
right. *Cash*, *Cover*, and *Annual* crops end at their harvest or
termination; *Perennial* (and *Woody perennial*) stands continue through
harvests until a termination; *Fallow* marks bare-fallow periods. Categories
are matched regardless of capitalization. A cover crop interseeded, broadcast,
or frost-seeded into a growing crop is entered with its own planting and
termination dates, and does not end at the companion crop's harvest.

## 2. Prepare the inputs

```{r prepare, eval = FALSE}
inputs <- prepare_shmi_inputs("my_SHMI_inputs.xlsx")   # your own workbook
```

With the installed example:

```{r prepare-example, message = FALSE}
inputs <- prepare_shmi_inputs(get_shmi_example(), verbose = FALSE)
names(inputs)
head(inputs$crop)
```

`prepare_shmi_inputs()`:

1. **Validates the workbook** with `validate_excel_input()` and stops with a
   list of problems if, for example, a required sheet or column is missing,
   a unit is not listed in `Mgt_Unit`, or a disturbance pass has no date.
2. **Builds crop episodes**: one row per species per continuous period of
   presence. Mixtures, relays, and intercrops are simply overlapping
   episodes. Harvest rows without a plant date are attached to a standing
   perennial as cuttings.
3. **Fills in missing dates**, following fixed rules. A crop with no plant
   date starts where the previous crop ended, or at the start of the window.
   A crop with no end date ends at the next planting, or at the end of the
   window, and earlier if intensive tillage is recorded after its planting
   (see *Crops without an end date* below).
4. **Sets the evaluation window** from 1 January of the first year with a
   record to 31 December of the last, so bare periods before the first and
   after the last recorded event count as bare.

It returns a list:

| Element | Contents |
|---|---|
| `rot_bounds` | Rotation start and end for each unit |
| `mgt` | Management-unit details |
| `crop` | Species episodes, with `crop_start` and `crop_end` |
| `dist`, `amend`, `animal` | Disturbance, amendment, and animal events |
| `yield`, `n_rate` | Yields and nitrogen rates, if requested |
| `assumptions` | Every date that was filled in, and every value to review |

### Reviewing assumptions

Management records are rarely complete, particularly at the start and end of
a record kept by calendar year. Rather than filling gaps silently,
`prepare_shmi_inputs()` records each assumption:

```{r assumptions}
table(inputs$assumptions$type)

# Items flagged for review, such as a perennial species entered as an annual
subset(inputs$assumptions, level == "check")
```

Rows with `level == "assumption"` are routine gap-filling. Rows with
`level == "check"` usually point to a data-entry issue worth correcting in
the workbook: a perennial species entered as an annual (or the reverse), an
unrecognized crop category or placeholder name, an annual crop lasting more
than 400 days, an imputed planting date (listed with the disturbance dates
that may be the real planting), a crop running to the end of the window
although tillage is recorded after planting, or a unit with no crops at all.
`?prepare_shmi_inputs` describes each type.

### Crops without an end date

A crop with no harvest or termination record first gets an imputed end: the
next planting or, if there is none, the end of the window. It then ends
earlier on the first day after planting with intensive tillage, meaning a day
whose passes sum to STIR ≥ 80, or whose EPA tillage intensity (mixed depth /
30 cm) is ≥ 0.252. Both thresholds are the lower bound of the
conventional-tillage class. Each such change is recorded as
`end_intensive_tillage`:

```{r tillage-end}
subset(inputs$assumptions, type == "end_intensive_tillage")
```

The scale is chosen from the data (`tillage_end = "auto"`: EPA when every pass
has a depth, otherwise STIR); set `tillage_end = "none"` to switch the rule
off.

### Choosing the evaluation window

By default the window runs over whole calendar years, from the first to the
last year with a record. To score a fixed period instead, give start and end
dates; events outside the window are dropped, crops and grazing periods are
clipped to it, and the window runs exactly between the two dates:

```{r window, message = FALSE}
inputs_2022_23 <- prepare_shmi_inputs(
  get_shmi_example(), verbose = FALSE,
  start_date_override = "2022-01-01",
  end_date_override   = "2023-12-31"
)
inputs_2022_23$rot_bounds
```

Other options include `exclude` (leave out some units), `end_at_sample_date`
(stop each unit at its soil sampling date, taken from the `MGT_sample_date`
column of `Mgt_Unit`), and `calc_yield` / `calc_n_rate` (also return yields
in kg/ha and annual nitrogen rates). Overrides are best placed on calendar-year
boundaries, because Organic Inputs and Inverse Disturbance are scored by
calendar year.

## 3. Add climate

Cover splits the year into a growing and a non-growing season for each
management unit. A month is in the growing season when its long-term mean
temperature is at least 5 °C and, unless the unit is irrigated, it is not dry
(precipitation in mm at least twice the mean temperature in °C).
Each unit's coordinates (`MGT_lat`, `MGT_lon`) and irrigation method
(`MGT_irr_cat`) are recorded on the Mgt_Unit tab. `get_shmi_climate()` takes
the monthly normals from WorldClim at those coordinates (it needs the
`geodata` and `terra` packages):

```{r climate-download, eval = FALSE}
climate <- get_shmi_climate(inputs)
```

The example's units are sprinkler-irrigated, so none of their months is
treated as too dry for growth:

```{r irrigated}
inputs$mgt[, c("MGT_combo", "MGT_irr_cat", "irrigated")]
```

Any source of monthly normals works, as long as the table has one row per
`MGT_combo` with `tavg_01` ... `tavg_12` (°C) and `prec_01` ... `prec_12`
(mm). Here, typical values for a temperate site are entered by hand:

```{r climate}
climate <- data.frame(MGT_combo = inputs$mgt$MGT_combo)
climate[sprintf("tavg_%02d", 1:12)] <- as.list(c(-5, -3, 3, 10, 16, 21, 24, 23, 18, 11, 4, -2))
climate[sprintf("prec_%02d", 1:12)] <- as.list(c(20, 25, 50, 70, 110, 115, 95, 85, 70, 50, 35, 25))
growing_months(climate[1, ])
```

## 4. Compute SHMI

```{r build, message = FALSE}
result <- build_shmi(inputs, climate)
result$indicator_df
```

`indicator_df` has one row per management unit, with `SHMI`, `Cover`,
`OrgInput`, `Diversity` and `InvDist`, the two parts of Cover
(`Cover_growing`, `Cover_nongrowing`), the average number of species per year
(`Richness`), and the unit's study, farm, field and treatment. The result also
records the weights, the tillage scale used (`dist_meth`), whether the scores
are official, the package version and a timestamp.

**Tillage.** `build_shmi()` computes tillage intensity as USDA's Tillage
Disturbance Index for Soil Carbon (T-DISC) does: per crop interval, from each
pass's mixing efficiency and depth. Each pass names its implement
(`SD_equip`), whose values come from T-DISC (`tdisc_implements()`,
`tdisc_mapping()`) or from a table you supply with `implements =`. `SD_mixeff`
and `SD_depth` (inches) are overrides: blank means the implement's values, and
either can be entered alone. STIR values go in `SD_stir` and are scored with
`dist_meth = "STIR"`, on the same crop intervals and windows.
The result also gives each unit's T-DISC tillage designation (NT, RT or CT).

## 5. Custom weights

For research or scenario analysis, the weights can be changed. They are
rescaled to sum to 1, and the scores are flagged as not official:

```{r custom, eval = FALSE}
result_equal <- build_shmi(inputs, climate,
                           weights = c(Cover = 0.25, OrgInput = 0.25,
                                       Diversity = 0.25, InvDist = 0.25))
shmi_weights()   # the official weights
```

## 6. How missing records are scored

SHMI treats a missing record as a practice that did not happen:

- **No disturbance records** means no tillage, so InvDist = 100. This is how
  continuous no-till is represented.
- **No crop on a given day** means no cover that day. A fallow reference site
  scores Cover = 0 and Diversity = 0. A unit with management records but no
  crop records is flagged (`no_crop_records`), so a forgotten crop sheet is
  not mistaken for a fallow.
- **No amendment or animal records** means no organic inputs, so
  OrgInput = 0.

Record every tillage pass, planting, harvest, and amendment for the period
being evaluated, so that gaps in the records are not mistaken for practices.

## 7. The sub-indices one at a time

Each sub-index can be computed directly from the prepared inputs:

```{r subindex-inputs}
compute_cover(inputs$crop, inputs$rot_bounds, climate)
compute_diversity(inputs$crop, inputs$rot_bounds)
compute_disturbance(inputs$dist, inputs$rot_bounds, inputs$crop, inputs$harvests)
compute_orginput(inputs$rot_bounds, inputs$amend, inputs$animal)
```

The functions also accept small tables built by hand, which is a quick way
to see how each one responds to a change in management. Here is one field
over two years: corn followed by a cereal rye cover crop, then soybean.

```{r toy-data}
crop <- data.frame(
  MGT_combo  = "field_1",
  CD_name    = c("Corn", "Rye", "Soybean"),
  crop_start = as.Date(c("2019-05-01", "2019-10-15", "2020-05-10")),
  crop_end   = as.Date(c("2019-09-30", "2020-04-20", "2020-09-25"))
)
rot_bounds <- data.frame(
  MGT_combo    = "field_1",
  rot_start    = as.Date("2019-01-01"),
  rot_end      = as.Date("2020-12-31"),
  rot_start_yr = 2019,
  rot_end_yr   = 2020
)
climate_1 <- climate[1, ]
climate_1$MGT_combo <- "field_1"
```

**Cover** is the share of days with a living crop in each season. The rye
keeps the soil covered through most of the non-growing season:

```{r toy-cover}
compute_cover(crop, rot_bounds, climate_1)
```

**Diversity** counts the species present each year: corn and rye in 2019,
rye and soybean in 2020, so two species a year. One species a year scores 0
and eight or more score 100:

```{r toy-diversity}
compute_diversity(crop, rot_bounds)
```

**Inverse disturbance** is computed as in USDA's T-DISC tool. Each cash crop
defines a crop interval, ending at its harvest; passes are grouped into
tillage windows around planting; and the interval is rated by its most
intense window. Here a heavy disk before corn and a planter pass each year
are given by implement name, so their mixing efficiency and depth come from
T-DISC (`tdisc_implements()`):

```{r toy-disturbance}
dist <- data.frame(
  MGT_combo = "field_1",
  SD_date   = as.Date(c("2019-04-20", "2019-05-01", "2020-05-10")),
  SD_equip  = c("HARROW, DISK, TANDEM, HEAVYDUTY", "PLANTER, REGULAR", "PLANTER, REGULAR")
)
harvests <- data.frame(
  MGT_combo = "field_1", CD_name = c("Corn", "Soybean"), CD_cat = "cash",
  harv_date = as.Date(c("2019-09-30", "2020-09-25"))
)
res <- compute_disturbance(dist, rot_bounds, crop, harvests, details = TRUE)
res
attr(res, "intervals")
```

The corn interval is rated 0.40 (the disk; conventional till), the soybean
interval only by its planter pass, and the score is their average weighted by
interval length. Values for any implement can be replaced with your own:

```{r toy-disturbance-override}
my_disk <- data.frame(implement = "HARROW, DISK, TANDEM, HEAVYDUTY",
                      mixing_efficiency = 0.6, depth_cm = 10)
compute_disturbance(dist, rot_bounds, crop, harvests, implements = my_disk)
```

**Organic inputs**: manure applied in one of the two years, and no animals:

```{r toy-orginput}
amend  <- data.frame(MGT_combo = "field_1",
                     SA_date = as.Date("2019-04-01"), SA_cat = "Organic")
animal <- data.frame(MGT_combo = character(),
                     AD_start_date = as.Date(character()),
                     AD_end_date = as.Date(character()))
compute_orginput(rot_bounds, amend, animal)
```

## 8. Plotting results

One unit, with a gauge for each sub-index and for overall SHMI:

```{r plot-gauge, fig.width = 8, fig.height = 4}
plot_shmi_gauge(result$indicator_df, row = 1)
```

Many units, overall SHMI sorted by score:

```{r plot-lollipop, fig.width = 7, fig.height = 4}
plot_shmi_lollipop(result$indicator_df)
```

Select a unit by name with `MGT_combo = "..."` instead of `row`.

## 9. Interpreting SHMI

Higher scores indicate management that is more supportive of soil health:
more of the year under living cover, more frequent organic inputs, more plant
species each year, and less tillage.

When comparing units, keep the weights in mind. A unit that scores 0 on one
sub-index cannot score above 100 × (1 − that sub-index's weight): a system
with no organic inputs, for example, can reach at most 68.3 however well it
does otherwise. Looking at the four sub-indices alongside SHMI shows which
practices drive a unit's score. SHMI is most informative for comparing
practices at the same site; a single score predicted for a new site is
typically within about 13 soil-health points.

## Further reading

- `?prepare_shmi_inputs`: workbook structure, crop-date rules, the
  assumptions table, and overrides.
- `?build_shmi` and `?shmi_weights`: the index and its official weights.
- `?compute_cover`, `?growing_months`, `?get_shmi_climate`: Cover's seasons
  and climate.
- `?compute_diversity`, `?compute_disturbance`, `?compute_orginput`: the other
  sub-indices.
- `?shmi_components`: all four sub-indices, for calibration or inspection.
- `NEWS.md`: what changed in each version.
