---
title: "SHMI Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SHMI Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



```{r setup}
library(SHMI)
```

# Introduction

The Soil Health Management Index (SHMI) provides a quantitative, rotation‑scale assessment of soil health management practices. It integrates four biologically grounded sub-indices:

- **Cover** — seasonal plant presence  
- **Diversity** — rotation-scale crop diversity (Hill numbers)  
- **Inverse Disturbance** — mechanistic mixing-efficiency × depth metric  
- **Organic Inputs** — amendments and animal presence  

This vignette provides a complete walk-through of the SHMI workflow using the functions in the SHMI R package.

---

## Quick Start {#quick-start}

## 📘 Minimal Example

A complete SHMI workflow in just a few lines

```{r}
library(SHMI)

# 1. Retrieve the sample Excel file
example <- get_shmi_example()

# 2. Prepare inputs (validates structure and expands events)
df <- prepare_shmi_inputs(example)

# 3. Compute sub-index scores and SHMI
result <- build_shmi(df)

head(result)

# 4. Plot the results
# Plot the first management unit
plot_shmi_gauge(result$indicator_df, row = 1)

# Plot all available management units
plot_shmi_lollipop(result$indicator_df)
```

---

## Detailed Instructions

# 1. Preparing Inputs

The SHMI workflow begins with a standardized Excel workbook containing:

- `Mgt_Unit`  
- `Crop_Diversity`  
- `Soil_Disturbance`  
- `Amendment_Diversity`  
- `Animal_Diversity`  

The function `prepare_shmi_inputs()`:

- reads and validates all sheets  
- harmonizes crop windows  
- constructs rotation bounds  
- builds daily grids  
- computes daily disturbance  
- assembles amendment and animal events  

```{r}
# Example (replace with your file path)
example_file <- SHMI::get_shmi_example()

inputs <- prepare_shmi_inputs(example_file, verbose = TRUE)
```

The returned object is a named list containing:

- `rot_bounds`  
- `crop_harmonized`  
- `daily`  
- `daily_dist`  
- `amend`  
- `animal`  
- and additional metadata  

These objects feed directly into `build_shmi()`.

---

# 2. Computing SHMI

Once inputs are prepared, computing SHMI is straightforward:

```{r}
shmi <- build_shmi(inputs)

shmi$indicator_df
```

The output includes:

- **Cover**  
- **Diversity**  
- **InvDist**  
- **Animals**  
- **SHMI** (weighted composite)  

as well as:

- `settings_used`  
- `expert_mode`  
- `timestamp`  
- `shmi_version`  

---

# 3. Expert Mode

By default, SHMI uses the official national settings (locked mode). To explore scenarios or conduct research, you may override settings:

```{r}

custom <- list(
    # cover
    w_winter = 0.25,
    w_spring = 0.25,
    w_summer = 0.25,
    w_fall   = 0.25,

    # diversity
    hill      = 2,
    max_div   = 8,

    # SHMI weights
    w_cover    = 0.25,
    w_div      = 0.25,
    w_dist     = 0.25,
    w_ani      = 0.25
)

shmi_exp <- build_shmi(inputs, settings = custom, expert_mode = TRUE)
```

When `expert_mode = TRUE`, SHMI values are not comparable to the national SHMI scale.

---

# 4. Sub-Index Details

This section demonstrates how each sub-index is computed individually.

## 4.1 Cover

```{r}
compute_cover(inputs$daily, inputs$rot_bounds)
```

Cover is based on seasonal plant-days, normalized by rotation length and weighted by seasonal importance.

---

## 4.2 Diversity

```{r}
compute_diversity(inputs$crop_harmonized, inputs$daily)
```

Diversity uses Hill-number entropy metrics at the rotation scale, with mixture expansion and species-level plant-day aggregation.

---

## 4.3 Inverse Disturbance

```{r}
compute_disturbance(inputs$daily_dist, inputs$rot_bounds)
```

Disturbance is computed using a mechanistic mixing-efficiency × depth model:

- cumulative mechanical energy  
- profile penetration (Tₜ)  
- annual aggregation  
- rotation averaging  
- scaling to 0–100  

---

## 4.4 Organic Inputs

```{r}
compute_orginput(inputs$rot_bounds, inputs$amend, inputs$animal)
```

Organic inputs combine amendment and animal events using user-defined weights.

---

# 5. Interpreting SHMI

The final SHMI score reflects:

- year-round cover  
- crop diversity  
- reduced soil disturbance  
- organic inputs  

Higher SHMI values indicate management systems that are more supportive of soil health.

---

# Conclusion

The SHMI package provides a fast, reproducible, and biologically grounded framework for quantifying soil health management across diverse systems.

For more details, see:

- `?prepare_shmi_inputs`  
- `?build_shmi`  
- `?compute_cover`  
- `?compute_diversity`  
- `?compute_disturbance`  
- `?compute_orginput`
