---
title: "SHMI Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SHMI Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  show_source: false
---

## Introduction

The Soil Health Management Index (SHMI) provides a quantitative, rotation‑scale assessment of soil‑health management practices. It integrates four biologically grounded sub‑indices:

- **Cover** — seasonal plant presence  
- **Diversity** — rotation‑scale crop diversity (Hill numbers)  
- **Inverse Disturbance** — mechanistic mixing‑efficiency × depth metric  
- **Organic Inputs** — organic amendments and animal presence  

This vignette walks through the **standard SHMI workflow** using the official Excel template.

---

# **1. Preparing Inputs**

SHMI begins with a standardized Excel workbook containing the following sheets:

- `Mgt_Unit`  
- `Crop_Diversity`  
- `Soil_Disturbance`  
- `Amendment_Diversity`  
- `Animal_Diversity`  

Download the blank template:

```{r, eval = FALSE}
library(SHMI)

# Save the template to a specific file path
template_file <- "SHMI_template.xlsx"
download_shmi_template(path = template_file)
```

Then:
1. Open **SHMI_template.xlsx** in Excel  
2. Enter your management data into each sheet  
3. Save the completed file (e.g., `"my_SHMI_inputs.xlsx"`)

Prepare the inputs:

```{r, eval = FALSE}
inputs <- prepare_shmi_inputs("my_SHMI_inputs.xlsx")
```

`prepare_shmi_inputs()`:

- validates required sheets and columns  
- harmonizes crop windows  
- constructs rotation bounds  
- builds daily grids  
- computes daily disturbance  
- assembles amendment and animal events  

The returned object is a named list containing:

- `rot_bounds`  
- `crop_harmonized`  
- `daily`  
- `daily_dist`  
- `amend`  
- `animal`  
- metadata  

These feed directly into `build_shmi()`.

---

# **2. Computing SHMI**

Once inputs are prepared:

```{r, eval = FALSE}
result <- build_shmi(inputs)
result$indicator_df
```

The output includes:

- **Cover**  
- **Diversity**  
- **InvDist**  
- **OrgInputs**  
- **SHMI** (weighted composite)

Plus metadata:

- `settings_used`  
- `expert_mode`  
- `timestamp`  
- `shmi_version`  

---

# **3. Expert Mode**

By default, SHMI uses the official national settings (locked mode).  
To explore scenarios or conduct research, you may override settings:

```{r, eval = FALSE}
custom <- list(
  # cover
  w_winter = 0.25,
  w_spring = 0.25,
  w_summer = 0.25,
  w_fall   = 0.25,

  # diversity
  hill    = 2,
  max_div = 8,

  # SHMI pillar weights
  w_cover = 0.25,
  w_div   = 0.25,
  w_dist  = 0.25,
  w_ani   = 0.25
)

shmi_exp <- build_shmi(inputs, settings = custom, expert_mode = TRUE)
```

When `expert_mode = TRUE`, SHMI values are **not comparable** to the national scale.

---

# **4. Sub‑Index Details**

You may compute sub‑indices individually:

### Cover
```{r, eval = FALSE}
compute_cover(inputs$daily, inputs$rot_bounds)
```

### Diversity
```{r, eval = FALSE}
compute_diversity(inputs$crop_harmonized, inputs$daily)
```

### Inverse Disturbance
```{r, eval = FALSE}
compute_disturbance(inputs$daily_dist, inputs$rot_bounds)
```

### Organic Inputs
```{r, eval = FALSE}
compute_orginput(inputs$rot_bounds, inputs$amend, inputs$animal)
```

---

# **5. Interpreting SHMI**

The final SHMI score reflects:

- year‑round cover  
- crop diversity  
- reduced soil disturbance  
- organic inputs  

Higher SHMI values indicate management systems that are more supportive of soil health.

---

# Conclusion

The SHMI package provides a fast, reproducible, and biologically grounded framework for quantifying soil‑health management across diverse systems.

For more details:

- `?prepare_shmi_inputs`  
- `?build_shmi`  
- `?compute_cover`  
- `?compute_diversity`  
- `?compute_disturbance`  
- `?compute_orginput`
